Online Multiclass Linear Classification with Bandit Feedback

### 2018-11-19 arxiv-version/main.tex
Add proof for the strongly linearly separable case

% LaTeX packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fullpage}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{natbib}
\usepackage{tikz}

@@

@@

The notion of linear separability with a margin is standard. It used in the
full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. For completeness, we state the
algorithm below. It is a folklore result that if a set of labeled examples is
separable with a margin $\gamma$ and the norm of the examples is bounded by $R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\left\lfloor
2(R/\gamma)^2 \right \rfloor$ mistakes. Another folklore result is that
\textsc{Multiclass Perceptron} is an optimal algorithm in the sense that any
deterministic algorithm makes $\left\lfloor 2(R/\gamma)^2 \right \rfloor$
mistakes in the worst case. For completeness, we give proofs  in
Section~\ref{section:multiclass-perceptron-proofs} for both of these results.

@@

\begin{theorem}
Let $(V, \ip{\cdot}{\cdot})$ be any inner product space, let $K$ be a positive
integer, $\gamma$ a positive real number, $R$ be a non-negative real number
If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled examples
in $V \times \{1,2,\dots,K\}$ such that the examples are strongly separable with
margin $\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ then
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes at most $K (R/\gamma)^2$ mistakes.
\end{theorem}

\begin{proof}
TODO
\end{proof}

@@

\begin{theorem}[Mistake upper bound]
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, let $\gamma$ be a positive real number and let $R$ be a non-negative real
number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\lfloor
2(R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

@@

\begin{algorithm}[h]
\caption{\textsc{Full-Information Multiclass Classification}
\label{algorithm:mutliclass-classification}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Predict $\widehat y_t \in \{1,2,\dots,K\}$}
\STATE{Observe $y_t \in \{1,2,\dots,K\}$}
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

\begin{theorem}[Mistake lower bound]
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any deterministic algorithm for the
\textsc{Full-Information Multiclass Classification} problem there exists an
inner product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
and the algorithm makes at least $\lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

>>>>

% LaTeX packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fullpage}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{natbib}
\usepackage{thmtools}
\usepackage{tikz}

@@

The notion of linear separability with a margin is standard. It used in the
full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. For completeness, we state the
algorithm in Appendix~\ref{section:multiclass-perceptron-proofs}. It is a
folklore result that if a set of labeled examples is separable with a margin
$\gamma$ and the norm of the examples is bounded by $R$ then \textsc{Multiclass
Perceptron} algorithm makes at most $\left\lfloor 2(R/\gamma)^2 \right \rfloor$
mistakes. Another folklore result is that \textsc{Multiclass Perceptron} is
essentially optimal in the sense that any deterministic algorithm must make
$\left\lfloor (R/\gamma)^2 \right \rfloor$ mistakes in the worst case. For
completeness, we give proofs of both of these results in
Section~\ref{section:multiclass-perceptron-proofs}.

The notion of strong linear separability is new.

@@

@@

\begin{theorem}
Let $(V, \ip{\cdot}{\cdot})$ be any inner product space, let $K$ be a positive
integer, $\gamma$ a positive real number, $R$ be a non-negative real number If
$(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled examples in
$V \times \{1,2,\dots,K\}$ such that the examples are strongly separable with
margin $\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ then the
expected number of mistakes
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes is at most $K(K-1) \lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

\begin{proof}
Let $M = \sum_{t=1}^T z_t$ be the number of mistakes the algorithm makes. Let $A
= \sum_{t ~:~ S_t \neq \emptyset} z_t$ be the number of mistakes in the rounds
when $S_t \neq \emptyset$ and let $B = \sum_{t ~:~ S_t = \emptyset} z_t$ be the
number of mistakes in the rounds when $S_t = \emptyset$. Obviously, $M = A + B$.
We upper bound $A$ and $B$ separately.

Let $C$ be the number of times line 16 gets executed. Let $U$ be the number of
times line 16 or line 21 get executed. In other words, $U$ the number of times
the weight vectors are updated. Clearly, $U = A + C$. We know that $U \le
\frac{K} \lfloor (R/\gamma)^2 \rfloor$.

The key observation is that $\Exp[B] = (K-1) \Exp[C]$ since there the $1/K$
probability that the algorithm guesses the correct label correctly if $S_t =
\emptyset$. Putting all the information together, we get that
\begin{align*}
\Exp[M]
& = \Exp[A] + \Exp[B] \\
& = \Exp[A] + (K-1) \Exp[C]  \\
& \le (K-1) \Exp[A + C] \\
& = (K-1) \Exp[U]  \\
& \le K(K-1) \lfloor (R/\gamma)^2 \rfloor \; .
\end{align*}
\end{proof}

@@

\textsc{Multiclass Perceptron} is an algorithm for \textsc{Online Multiclass
Classification}. Both the protocol for the problem and the algorithm are stated
below. The algorithm assumes that the feature vectors come from an inner product
space $(V, \ip{\cdot}{\cdot})$.

Two results are folklore. The first result is
\autoref{theorem:mutliclass-perceptron-mistake-upper-bound} which states that if
examples are linearly separable with margin $\gamma$ and examples have norm
at most $R$ then the algorithm makes at most $\lfloor 2 (R/\gamma)^2 \rfloor$
mistakes. The second result is
\autoref{theorem:online-multiclass-classification-mistake-lower-bound} which
states that under the same assumptions as in
\autoref{theorem:online-multiclass-classification-mistake-lower-bound}
\emph{any} deterministic algorithm for \textsc{Online Multiclass Classification}
must make at least $\lfloor (R/\gamma)^2 \rfloor$ mistakes in the worst case.

\begin{algorithm}[h]
\caption{\textsc{Online Multiclass Classification}
\label{algorithm:mutliclass-classification}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Predict $\widehat y_t \in \{1,2,\dots,K\}$}
\STATE{Observe $y_t \in \{1,2,\dots,K\}$}
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

\begin{theorem}[Mistake upper bound]
\label{theorem:mutliclass-perceptron-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, let $\gamma$ be a positive real number and let $R$ be a non-negative real
number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\lfloor
2(R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:online-multiclass-classification-mistake-lower-bound}
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any deterministic algorithm for the
\textsc{Full-Information Multiclass Classification} problem there exists an
inner product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
and the algorithm makes at least $\lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}
